---
title: "02_sequence_analysis.Rmd"
author: "Scott Klasek"
date: "6/7/2020"
output: github_document
---

## 0) load libraries and import phyloseq object:

```{r libraries}
library(tidyr)
library(plyr)
library(dplyr)
library(ggplot2)
library(phyloseq)
library(DESeq2)
library(vegan)
library(metagenomeSeq)
library(patchwork)
library(RColorBrewer)
library(here)
here()
ps4 <- readRDS(file = "ps4")
```

## 1) Bar plots showing high-level (class) community composition of all samples: 

```{r}
ps4.ra <- transform_sample_counts(ps4, function(OTU) OTU/sum(OTU)) # transform read counts to relative abundances

# graph sequences belonging to most abundant classes (as done with the OTU dataset) 
classes.ps4 <- unique(tax_table(ps4)[,3]) # obtains the 122 unique class names
classes.ps4 <- classes.ps4[!is.na(classes.ps4)] # removes the NA (121 unique classes)
mx.asv.ra <- as.data.frame(otu_table(ps4.ra)) # creates a matrix from the phyloseq object
class.ra <- vector("list",0) # define vector
for (i in classes.ps4) {class.ra[[i]] <- sum(mx.asv.ra[,which(tax_table(ps4.ra)[,3]==i)])/44} # calculate average % abund of each class
class.output <- cbind(class.ra)[,1] 
class.output.n <- as.numeric(cbind(class.ra)[,1]) # convert output to numeric
class.output.df <- class.output.df[order(-class.output.df$class.output.n),] # sort by pct abundance in descending order
top.c <- as.character(class.output.df[1:17,1]) # the most abundant 17 classes are each > 1% relabund, selecting these
sum(class.output.df[1:17,2]) # 77% of reads are from these top 17 classes
sum(class.output.df[1:10,2]) # 68% of reads are from these top 10 classes
ps4.ra.top.c <- subset_taxa(ps4.ra, Class=="Deltaproteobacteria" | Class=="Gammaproteobacteria" | Class=="Anaerolineae" | Class=="Alphaproteobacteria" | Class==  "Ignavibacteria" | Class=="Nitrospira" | Class=="Thermodesulfovibrionia" | Class=="Bacteroidia" | Class=="Campylobacteria" | Class=="Phycisphaerae") # top classes. just setting Class==top.c doesn't work and eliminates a lot of ASVs for a reason I don't understand

# testing a few bar plots before moving to the final one:
top.c.bp <- plot_bar(ps4.ra.top.c, fill="Class")+
  scale_y_continuous(limits = c(0,0.9)) # barplot showing ASVs belonging to classes > 2% abundance
ps4.ra.top300.names <- names(sort(taxa_sums(ps4.ra), decreasing=TRUE))[1:300]
ps4.ra.top300 <- prune_taxa(ps4.ra.top300.names, ps4.ra)
top300bp <- plot_bar(ps4.ra.top300, fill="Class") # barplot showing most abundant 300 ASVs regardless of class
# top300bp is more sparse and there are more classes, BUT it shows there are a few class==NA ASVs highly abundant in South core tsunami deposits. they appear to be Zixibacteria

# Now the final barplot
# subset four pseudo-panels (North/South x Scanned/Unscanned) that are each individual plots, but arrange them together
sample_data(ps4.ra.top.c)$label <- (paste(sample_data(ps4.ra.top.c)$depth_cm, "cm,", sample_data(ps4.ra.top.c)$storage_days, "d")) # add a character vector to plot by

ps4.ra.top.c.north.s <- subset_samples(ps4.ra.top.c, site=="North" & scan=="Scanned") # subset the North samples (scanned and unscanned)
ps4.ra.top.c.north.u <- subset_samples(ps4.ra.top.c, site=="North" & scan=="Unscanned") 
ps4.ra.top.c.south.s <- subset_samples(ps4.ra.top.c, site=="South" & scan=="Scanned") # subset the South samples (scanned and unscanned)
ps4.ra.top.c.south.u <- subset_samples(ps4.ra.top.c, site=="South" & scan=="Unscanned") 

gg.bar.top.c.north.s <- plot_bar(ps4.ra.top.c.north.s, x="label", fill="Class") +
  geom_rect(aes(xmin=3.5, xmax=6.5, ymin=-Inf, ymax=Inf), fill="gray80", alpha=0.01) +
  geom_bar(stat="identity", position="stack") + 
  scale_fill_brewer(palette = "Set3") +
  scale_x_discrete("") +
  facet_grid(~scan, scales = "free", space = "free") +
  ggtitle("North Site") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        strip.text.x = element_text(size=11))

gg.bar.top.c.north.u <- plot_bar(ps4.ra.top.c.north.u, x="label", fill="Class") +
  geom_rect(aes(xmin=5.5, xmax=9.5, ymin=-Inf, ymax=Inf), fill="gray80", alpha=0.01) +
  geom_bar(stat="identity", position="stack") + 
  scale_fill_brewer(palette = "Set3") +
  scale_x_discrete("") +
  scale_y_continuous("") +
  facet_grid(~scan, scales = "free", space = "free") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1), axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), strip.text.x = element_text(size=11))

gg.bar.top.c.south.s <- plot_bar(ps4.ra.top.c.south.s, x="label", fill="Class") +
  geom_rect(aes(xmin=2.5, xmax=6.5, ymin=-Inf, ymax=Inf), fill="gray80", alpha=0.01) +
  geom_bar(stat="identity", position="stack") + 
  scale_fill_brewer(palette = "Set3") +
  scale_x_discrete("Sample") +
  facet_grid(~scan, scales = "free", space = "free") +
  ggtitle("South Site") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        strip.text.x = element_text(size=11))

gg.bar.top.c.south.u <- plot_bar(ps4.ra.top.c.south.u, x="label", fill="Class") +
  geom_rect(aes(xmin=5.5, xmax=10.5, ymin=-Inf, ymax=Inf), fill="gray80", alpha=0.01) +
  geom_bar(stat="identity", position="stack") + 
  scale_fill_brewer(palette = "Set3") +
  scale_x_discrete("Sample") +
  scale_y_continuous("") +
  facet_grid(~scan, scales = "free", space = "free") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1), axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), strip.text.x = element_text(size=11))

barplot <- (gg.bar.top.c.north.s + gg.bar.top.c.north.u) / (gg.bar.top.c.south.s + gg.bar.top.c.south.u) + plot_layout(guides="collect")
barplot
```

## 2) Alpha diversity analysis of all samples

```{r}
paircode <- read.csv(file="paircode.csv") # import the file showing which samples are paired with one another
sample_data(ps4) <- data.frame(sample_data(ps4), paircode) # add this info to the ps object

plot_richness(ps4, x="scan", measures = c("Observed", "Shannon")) # simple plot

alphadiv <- estimate_richness(ps4, measures = c("Observed", "Shannon")) # calculate alpha diversity
alphadiv <- data.frame(alphadiv, sample_data(ps4)) # write metadata into an alpha diversity data frame

# Is there a difference in alpha diversity between scanned and unscanned samples?
t.test(alphadiv[which(alphadiv$scan=="Unscanned"),]$Observed, alphadiv[which(alphadiv$scan=="Scanned"),]$Observed) # observed richness, no
t.test(alphadiv[which(alphadiv$scan=="Unscanned"),]$Shannon, alphadiv[which(alphadiv$scan=="Scanned"),]$Shannon) # shannon, no

# Alpha diversity should be analyzed by pairing scanned/unscanned samples:
df.ps4 <- data.frame(sample_data(ps4), alphadiv[,1:3]) # combine sample data and alpha diversity into a df
df.ps4.p <- df.ps4 %>% filter(is.paired=="y") # filter only the paired samples
df.ps4.p <- df.ps4.p[order(df.ps4.p$paircode),] # order the dataframe by their paircode
t.test(df.ps4.p[which(df.ps4.p$scan=="Unscanned"),]$Observed, df.ps4.p[which(df.ps4.p$scan=="Scanned"),]$Observed, paired = TRUE) # t = -1.686, df = 10, p-value = 0.1227
t.test(df.ps4.p[which(df.ps4.p$scan=="Unscanned"),]$Shannon, df.ps4.p[which(df.ps4.p$scan=="Scanned"),]$Shannon, paired = TRUE) # t = -2.0386, df = 10, p-value = 0.06881

# plot alpha diversity over time
gg.alphadiv.obs <- ggplot(alphadiv, aes(storage_days, Observed))+
  geom_point()+
  facet_grid(~site)
gg.alphadiv.obs # no huge changes in observed ASV richness

gg.alphadiv.sha <- ggplot(alphadiv, aes(storage_days, Shannon))+
  geom_point()+
  facet_grid(~site)
gg.alphadiv.sha # no huge changes in Shannon diversity over time

# alpha diversity by site (north/south)
t.test(alphadiv[which(alphadiv$site=="North"),]$Observed, alphadiv[which(alphadiv$site=="South"),]$Observed) # not in observed richness
t.test(alphadiv[which(alphadiv$site=="North"),]$Shannon, alphadiv[which(alphadiv$site=="South"),]$Shannon) # not in Shannon

# alpha diversity by tsunami layer
t.test(alphadiv[which(alphadiv$sedtype=="Estuary Sediment"),]$Observed, alphadiv[which(alphadiv$sedtype=="Tsunami Deposit"),]$Observed) # not in observed richness
t.test(alphadiv[which(alphadiv$sedtype=="Estuary Sediment"),]$Shannon, alphadiv[which(alphadiv$sedtype=="Tsunami Deposit"),]$Shannon) # not in Shannon
```


## 3) Normalizations and ordinations

```{r}
## First, transform data, try Hellinger transformation
otu_table(ps4)[1:5,1:5]
otu.ps4.hell <- otu_table(decostand(otu_table(ps4), method = "hellinger"), taxa_are_rows=FALSE)
ps4.hel <- phyloseq(tax_table(ps4),
                    sample_data(ps4),
                    otu_table(otu.ps4.hell),
                    phy_tree(ps4),
                    refseq(ps4))

# Another transformation: cumulative sum squares
write.csv(otu_table(ps4),"ps4.otu.table")
ps4.otu <- read.csv(file = "ps4.otu.table", header = TRUE) # Why did I write a .csv just to reimport it? https://github.com/HCBravoLab/metagenomeSeq/issues/76
rownames(ps4.otu) <- ps4.otu[,1]
ps4.otu <- ps4.otu[,-1]
ps4.otu <- t(ps4.otu)
ps4.otu.MR <- newMRexperiment(ps4.otu)
ps4.otu.css <- cumNorm(ps4.otu.MR) 
ps4.otu.css <- MRcounts(ps4.otu.css, norm = T)
ps4.css <- phyloseq(tax_table(ps4),
                    sample_data(ps4),
                    otu_table(ps4.otu.css, taxa_are_rows=TRUE),
                    phy_tree(ps4),
                    refseq(ps4))

## Now generate distance matrices:
# binary jaccard here
dm.ps4.hel.jac <- phyloseq::distance(ps4.hel, method = "jaccard", binary = TRUE) 
dm.ps4.css.jac <- phyloseq::distance(ps4.css, method = "jaccard", binary = TRUE)
# weighted unifrac here
dm.ps4.hel.wu <- phyloseq::distance(ps4.hel, method = "wunifrac") 
dm.ps4.css.wu <- phyloseq::distance(ps4.css, method = "wunifrac")

## Make ordinations
ord.ps4.hel.nmds.bc <- ordinate(ps4.hel, "NMDS", "bray") # no need to specify a distance matrix with bray-curtis
ord.ps4.css.nmds.bc <- ordinate(ps4.css, "NMDS", "bray")
ord.ps4.hel.pcoa.bc <- ordinate(ps4.hel, "PCoA", "bray") 
ord.ps4.css.pcoa.bc <- ordinate(ps4.css, "PCoA", "bray")
ord.ps4.css.pcoa.bjac <- ordinate(ps4, method="PCoA", distance=dm.ps4.css.jac) # don't need to transform with binary jaccard
ord.ps4.css.pcoa.wu <- ordinate(ps4, method="PCoA", distance=dm.ps4.css.wu)
ord.ps4.hel.pcoa.wu <- ordinate(ps4, method="PCoA", distance=dm.ps4.hel.wu)

## Generate a few rough-draft ordination plots:
ord1 <- plot_ordination(ps4.hel, ord.ps4.hel.nmds.bc, color="site", title="NMDS Bray-Curtis hellinger, stress=0.067 but no convergence")
ord2 <- plot_ordination(ps4.css, ord.ps4.css.nmds.bc, color="site", title="NMDS Bray-Curtis CSS, stress=0.0747")
ord3 <- plot_ordination(ps4.hel, ord.ps4.hel.pcoa.bc, color="site", title="PCoA Bray-Curtis hellinger")
ord4 <- plot_ordination(ps4.css, ord.ps4.css.pcoa.bc, color="site", title="PCoA Bray-Curtis CSS")
# PCoAs seem to show better clustering than NMDS. Though minimal difference in transformation types, CSS shows just a little more variation than hellinger.
ord5 <- plot_ordination(ps4.css, ord.ps4.css.pcoa.bjac, color="site", title="PCoA Binary Jaccard") # binary jaccard shows least variance 
ord6 <- plot_ordination(ps4.css, ord.ps4.css.pcoa.wu, color="scan", shape="site", title="PCoA weighted unifrac, CSS") # 54 by 10%
ord7 <- plot_ordination(ps4.hel, ord.ps4.hel.pcoa.wu, color="site", title="PCoA weighted unifrac, hellinger")
# weighted unifrac shows by far the most variance. hellinger shows slightly more variation but less separation in N samples

# Split up the CSS-transformed phyloseq objects by North/South
ps4.css.north <- subset_samples(ps4.css, site=="North")
ps4.css.south <- subset_samples(ps4.css, site=="South")

# Add in the above/below/within tsunami deposit. Note that tsunami deposit depths are different in different cores
sample_data(ps4.css.north)$position <- ifelse(sample_data(ps4.css.north)$sedtype=="Tsunami Deposit", "Within",
  ifelse(sample_data(ps4.css.north)$sedtype=="Estuary Sediment" & sample_data(ps4.css.north)$depth_cm < 60, "Above",
  ifelse(sample_data(ps4.css.north)$sedtype=="Estuary Sediment" & sample_data(ps4.css.north)$depth_cm > 62, "Below", NA)))
sample_data(ps4.css.south)$position <- ifelse(sample_data(ps4.css.south)$sedtype=="Tsunami Deposit", "Within",
  ifelse(sample_data(ps4.css.south)$sedtype=="Estuary Sediment" & sample_data(ps4.css.south)$depth_cm < 69, "Above",
  ifelse(sample_data(ps4.css.south)$sedtype=="Estuary Sediment" & sample_data(ps4.css.south)$depth_cm > 70, "Below", NA)))

# Ordinate North/South separately, and label by above/within/below deposit (or sedtype) and add label=depth_cm. Maybe also storage time?
dm.ps4.css.north.wu <- phyloseq::distance(ps4.css.north, method = "wunifrac")
dm.ps4.css.south.wu <- phyloseq::distance(ps4.css.south, method = "wunifrac")
ord.ps4.css.north.pcoa.wu <- ordinate(ps4.css.north, method="PCoA", distance=dm.ps4.css.north.wu)
ord.ps4.css.south.pcoa.wu <- ordinate(ps4.css.south, method="PCoA", distance=dm.ps4.css.south.wu)
ord8 <- plot_ordination(ps4.css.north, ord.ps4.css.north.pcoa.wu, color="storage_days", label="depth_cm", title="North- PCoA weighted unifrac, CSS")
ord9 <- plot_ordination(ps4.css.south, ord.ps4.css.south.pcoa.wu, color="storage_days", label="depth_cm", title="South- PCoA weighted unifrac, CSS")
ord10 <- plot_ordination(ps4.css.north, ord.ps4.css.north.pcoa.wu, color="position", label="depth_cm", title="North- PCoA weighted unifrac, CSS")
ord11 <- plot_ordination(ps4.css.south, ord.ps4.css.south.pcoa.wu, color="position", label="depth_cm", title="South- PCoA weighted unifrac, CSS")

# final ordination plots
gg.ord.all <- plot_ordination(ps4.css, ord.ps4.css.pcoa.wu, color="scan", shape="site", title="PCoA weighted unifrac, CSS") #
gg.ord.all

gg.ord.storage <- ord8 + ord9 + plot_layout(guides="collect")
gg.ord.storage

gg.ord.sites <- ord10 + ord11 + plot_layout(guides="collect")
gg.ord.sites

```

1 of 2 NMDS converged, and the PCoAs captured under 50% of the variation. PCoAs seem to show better clustering though
Little difference in the transformations (Hellinger or CSS) but CSS shows just a little more variation























